---
title: "Final Code Hazmat Investigations"
name: Katherine Oung
date: "2023-07-22"
output:
  pdf_document: default
  html_document: default
---

```{r import libraries}
library(tidyverse)
library(lubridate)
library(plotly)
library(dplyr)
library(readxl)
library(scales)
library(janitor)
options(scipen = 9999)

```

#################################################################
Setting Up the Data
- reading data in
- fixing date information
- removing rows so you only see incidents (accounting for "multiple rows per incident")
- adjust incidents 
#################################################################

The first thing to do is to import your two datasets of hazmat incidents. The first is a detailed incident report of all hazmat incidents in the whole country. The second is a detailed incident report of hazmat incidents in the five-state region (IN, IL, OH, MI, KY)
```{r}
hazmat_allstates <- read.csv('AllStates_Incident Detailed Report All fields included in Form 5800 (2).csv')
hazmat_region_full <- read.csv("Hazmat-FULLREPORT-Incident-Detailed-Report-All-fields-included-in-Form-5800.csv")
```

Fix dates
- Region data is in Month-date-year format
- Country data is in Year-month-date format
```{r}
#fix date column
hazmat_allstates <- hazmat_allstates %>%
  mutate(Date.Of.Incident = ymd(Date.Of.Incident)) %>%
  mutate(year = year(Date.Of.Incident)) %>%
  mutate(year = as.integer(year))

#fix date column
hazmat_region_full <- hazmat_region_full %>%
  mutate(Date.Of.Incident = mdy(Date.Of.Incident)) %>%
  mutate(year = year(Date.Of.Incident)) %>%
  mutate(year = as.integer(year))
```

Make into incidents
```{r}
#all states incidents
hazmat_allstates_incidents <- hazmat_allstates %>%
  select(Report.Number, Multiple.Rows.Per.Incident, Carrier.Reporter.Name, Mode.Of.Transportation, Incident.County, Environmental.Damage.Result, Water.Sewer.Result.Ind, Incident.City, Incident.State, Date.Of.Incident, Serious.Incident.Ind, Transportation.Phase, Total.Amount.Of.Damages, Total.Hazmat.Injuries, Total.Hazmat.Fatalities, year)

hazmat_allstates_incidents <- distinct(hazmat_allstates_incidents) %>%
    mutate(countystate = paste(Incident.County, " COUNTY, ", Incident.State, sep = ""))

#region incidents
hazmat_region_full_incidents <- hazmat_region_full %>%
  select(Report.Number, Multiple.Rows.Per.Incident, Carrier.Reporter.Name, Mode.Of.Transportation, Incident.County, Environmental.Damage.Result, Water.Sewer.Result.Ind, Incident.City, Incident.State, Date.Of.Incident, Serious.Incident.Ind, Transportation.Phase, Total.Amount.Of.Damages, Total.Hazmat.Injuries, Total.Hazmat.Fatalities, year)

hazmat_region_full_incidents <- distinct(hazmat_region_full_incidents) %>%
    mutate(countystate = paste(Incident.County, " COUNTY, ", Incident.State, sep = ""))
```

Adjust for inflation
```{r adjust country values for inflation using BEA values}
#add fiscal year value
hazmat_allstates_incidents <- hazmat_allstates_incidents %>%
  mutate(Fiscal.Year = case_when(
    Date.Of.Incident <= as.Date('2013-09-30') ~ 2013,
    Date.Of.Incident <= as.Date('2014-09-30') ~ 2014,
    Date.Of.Incident <= as.Date('2015-09-30') ~ 2015,
    Date.Of.Incident <= as.Date('2016-09-30') ~ 2016,
    Date.Of.Incident <= as.Date('2017-09-30') ~ 2017,
    Date.Of.Incident <= as.Date('2018-09-30') ~ 2018,
    Date.Of.Incident <= as.Date('2019-09-30') ~ 2019,
    Date.Of.Incident <= as.Date('2020-09-30') ~ 2020,
    Date.Of.Incident <= as.Date('2021-09-30') ~ 2021,
    Date.Of.Incident <= as.Date('2022-09-30') ~ 2022,
    Date.Of.Incident <= as.Date('2023-09-30') ~ 2023,
  ))


#adjust for inflation using BEA values
adjust_values <- read.csv("../final_data_product/adjustforinflationvalues_BEA.csv")

hazmat_allstates_incidents <- left_join(hazmat_allstates_incidents, adjust_values, by = "Fiscal.Year")


hazmat_allstates_adjusted_incidents <- hazmat_allstates_incidents %>%
  mutate(Adjusted.Total.Damages = Total.Amount.Of.Damages * values)

hazmat_allstates_adjusted_incidents %>%
  filter(Incident.City == "EAST PALESTINE")

hazmat_allstates_incidents %>%
  filter(is.na(Total.Amount.Of.Damages))

hazmat_allstates_adjusted_incidents %>%
  arrange(-Adjusted.Total.Damages)

hazmat_allstates_adjusted_incidents %>%
  group_by(Mode.Of.Transportation) %>%
  count(year) %>%
  mutate(percentage = n / sum(n) * 100)
```



```{r adjust region values for inflation using BEA values}
#add fiscal year value
hazmat_region_full_incidents <- hazmat_region_full_incidents %>%
  mutate(Fiscal.Year = case_when(
    Date.Of.Incident <= as.Date('2013-09-30') ~ 2013,
    Date.Of.Incident <= as.Date('2014-09-30') ~ 2014,
    Date.Of.Incident <= as.Date('2015-09-30') ~ 2015,
    Date.Of.Incident <= as.Date('2016-09-30') ~ 2016,
    Date.Of.Incident <= as.Date('2017-09-30') ~ 2017,
    Date.Of.Incident <= as.Date('2018-09-30') ~ 2018,
    Date.Of.Incident <= as.Date('2019-09-30') ~ 2019,
    Date.Of.Incident <= as.Date('2020-09-30') ~ 2020,
    Date.Of.Incident <= as.Date('2021-09-30') ~ 2021,
    Date.Of.Incident <= as.Date('2022-09-30') ~ 2022,
    Date.Of.Incident <= as.Date('2023-09-30') ~ 2023,
  ))

#edit east palestine value!!!
hazmat_region_full_incidents <- hazmat_region_full_incidents %>%
    mutate(Total.Amount.Of.Damages = ifelse(Incident.City == "EAST PALESTINE", 58520000, Total.Amount.Of.Damages))
  

#adjust for inflation using BEA values
adjust_values <- read.csv("../final_data_product/adjustforinflationvalues_BEA.csv")
hazmat_region_full_incidents <- left_join(hazmat_region_full_incidents, adjust_values, by = "Fiscal.Year")


hazmat_region_adjusted_incidents <- hazmat_region_full_incidents %>%
  mutate(Adjusted.Total.Damages = Total.Amount.Of.Damages * values)

hazmat_region_adjusted_incidents %>%
  arrange(-Adjusted.Total.Damages)

```


```{r}
county_codes <- read.csv("../final_data_product/NCHSURCodes2013.csv")

county_codes <- county_codes %>%
  select(County.name, X2013.code, State) %>%
  mutate(countystate = paste(County.name, ", ", State, sep = ""))
  

county_codes$countystate <- toupper(county_codes$countystate)

hazmat_ten_full_codes <- left_join(hazmat_region_adjusted_incidents, county_codes, by = "countystate")

hazmat_ten_full_codes %>%
  filter(is.na(X2013.code))

hazmat_map_all <- hazmat_ten_full_codes %>%
    filter(!is.na(X2013.code))

```


#################################################################
Datapoints for Main Story
#################################################################

Proportion of incidents in the region
```{r}
hazmat_allstates_adjusted_incidents #210,518

hazmat_region_adjusted_incidents #46,049

hazmat_region_adjusted_incidents%>%
    filter(Incident.State == "IL" | Incident.State == "IN" | Incident.State == "OH" | Incident.State == "MI" | Incident.State == "KY") %>%
    count(Incident.State) %>%
  arrange(-n)

21.8%
```

Annual number of incidents in the region
```{r}
hazmat_region_adjusted_incidents %>%
  count(year)
```

Percent/number of incidents deemed serious
```{r}
hazmat_region_adjusted_incidents %>%
  count(Serious.Incident.Ind) %>%
  mutate(percentage = n / sum(n) *100)
```

Incidents by mode
```{r}
hazmat_region_adjusted_incidents %>%
  count(Mode.Of.Transportation) %>%
  mutate(percentage = n / sum(n) *100)
```

Number of fatalities
```{r}
hazmat_region_adjusted_incidents %>%
  summarize(sum = sum(Total.Hazmat.Fatalities))

hazmat_region_adjusted_incidents %>%
  summarize(sum = sum(Total.Hazmat.Injuries))
```

20 worst incidents — what companies - Norfolk Southern
```{r}
hazmat_region_adjusted_incidents %>%
  arrange(-Adjusted.Total.Damages) %>%
  select(Carrier.Reporter.Name,Mode.Of.Transportation, Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values)

7/20
```

Last year
```{r}
hazmat_region_adjusted_incidents %>%
  filter(year == 2022) 

hazmat_region_adjusted_incidents %>%
  filter(year == 2022) %>%
  filter(Serious.Incident.Ind == "Yes")

hazmat_allstates_adjusted_incidents %>%
    filter(Incident.State == "IL" | Incident.State == "IN" | Incident.State == "OH" | Incident.State == "MI" | Incident.State == "KY") %>%
    filter(year == 2022) %>%
    filter(Serious.Incident.Ind == "Yes")


```
City with the most incidents
```{r}
hazmat_region_adjusted_incidents %>%
  count(Incident.City) %>%
  arrange(-n)

hazmat_region_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  count(Incident.City) %>%
  arrange(-n)
```

#################################################################
Datapoints for Data Article
#################################################################
```{r}
hazmat_allstates_adjusted_incidents %>%
  filter(Incident.State == "IL" | Incident.State == "IN" | Incident.State == "OH" | Incident.State == "MI" | Incident.State == "KY") %>%
  count(year) #bye why is it diff
  

hazmat_region_adjusted_incidents %>%
    count(year)

hazmat_region_adjusted_incidents %>%
  filter(Multiple.Rows.Per.Incident == "Yes") %>%
  arrange(Date.Of.Incident)

hazmat_region_full %>%
  count(year)

hazmat_allstates %>%
  filter(Incident.State == "IL" | Incident.State == "IN" | Incident.State == "OH" | Incident.State == "MI" | Incident.State == "KY") %>%
  count(year)

hazmat_region_adjusted_incidents %>%
  summarize(sum(Adjusted.Total.Damages))
```
Over time
```{r}
hazmat_allstates_adjusted_incidents %>% #seemed like they took out highway accidents in between the time of me downloading the report lmfaooooo
  filter(Incident.State == "IL" | Incident.State == "IN" | Incident.State == "OH" | Incident.State == "MI" | Incident.State == "KY") %>%
  group_by(Mode.Of.Transportation) %>%
  count(year)
  

hazmat_region_adjusted_incidents %>%
  group_by(Mode.Of.Transportation) %>%
    count(year)

hazmat_region_full %>%
  count(year)

```

In transit
```{r}
hazmat_region_adjusted_incidents %>%
  count(Transportation.Phase) %>%
  mutate(percentage = n/sum(n))

hazmat_allstates_adjusted_incidents %>%
  count(Transportation.Phase) %>%
  mutate(percentage = n/sum(n))

```
```{r}
hazmat_region_adjusted_incidents %>%
  group_by(Transportation.Phase) %>%
  summarize(damages = sum(Adjusted.Total.Damages)) %>%
  mutate(percentage = damages/sum(damages))


hazmat_allstates_adjusted_incidents %>%
  group_by(Transportation.Phase) %>%
  summarize(damages = sum(Adjusted.Total.Damages)) %>%
  mutate(percentage = damages/sum(damages))

```
```{r in transit by mode}
hazmat_region_graph <- hazmat_region_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  group_by(year, Mode.Of.Transportation) %>%
  summarise(count = n()) 
  
hazmat_region_graph %>%
  ggplot(aes(x = year, y = count, fill = Mode.Of.Transportation)) +
  geom_col()
```

```{r}
hazmat_region_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  count(Serious.Incident.Ind)

```
Average damages in the country
```{r in the country}
hazmat_allstates_adjusted_incidents %>%
  group_by(Mode.Of.Transportation) %>%
  summarize(sum = sum(Adjusted.Total.Damages)) %>%
  arrange(-sum) %>%
  mutate(percentage = sum/sum(sum) *100)

hazmat_allstates_adjusted_incidents %>%  #3317.56		
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_allstates_adjusted_incidents %>% #72474.35
  filter(Mode.Of.Transportation == "Rail") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

####

hazmat_allstates_adjusted_incidents %>%  #10521.21
  filter(Transportation.Phase == "In Transit") %>%
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_allstates_adjusted_incidents %>% #73842
  filter(Transportation.Phase == "In Transit") %>%
  filter(Mode.Of.Transportation == "Rail") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_allstates_adjusted_incidents %>%  #3317.56		
  filter(Transportation.Phase == "In Transit") %>%
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))


hazmat_allstates_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  group_by(Mode.Of.Transportation) %>%
  summarize(sum = sum(Adjusted.Total.Damages)) %>%
  mutate(percentage = sum/sum(sum) *100)


hazmat_allstates_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  count(Mode.Of.Transportation)


hazmat_allstates_adjusted_incidents %>%  #3317.56		
  filter(Serious.Incident.Ind == "Yes") %>%
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_allstates_adjusted_incidents %>% #72474.35
  filter(Serious.Incident.Ind == "Yes") %>%
  filter(Mode.Of.Transportation == "Rail") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

72474.35/3317.56

hazmat_allstates_adjusted_incidents %>%  #3317.56		
  filter(Transportation.Phase == "Unloading") %>%
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_allstates_adjusted_incidents %>% #72474.35
  filter(Transportation.Phase == "Unloading") %>%
  filter(Mode.Of.Transportation == "Rail") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

```


```{r}
hazmat_region_full_incidents %>%
  filter(Environmental.Damage.Result == "Yes" | Water.Sewer.Result.Ind == "Yes")

hazmat_allstates_incidents %>%
  filter(Environmental.Damage.Result == "Yes" | Water.Sewer.Result.Ind == "Yes")

```

#################################################################
Map Of Incidents
#################################################################
```{r}
hazmat_map_all <- hazmat_map_all %>%
  group_by(Incident.State) %>%
  count(countystate) %>%
  arrange(-n)
```

#################################################################
Five worst incidents
#################################################################
```{r}
hazmat_region_adjusted_incidents %>%
  select(Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)
```


```{r}
hazmat_region_adjusted_incidents %>%
  filter(Incident.State == "IL") %>%
  select(Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)

hazmat_region_adjusted_incidents %>%
  filter(Incident.State == "IN") %>%
  select(Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)

hazmat_region_adjusted_incidents %>%
  filter(Incident.State == "OH") %>%
  select(Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)

hazmat_region_adjusted_incidents %>%
  filter(Incident.State == "KY") %>%
  select(Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)

hazmat_region_adjusted_incidents %>%
  filter(Incident.State == "MI") %>%
  select(Mode.Of.Transportation, Date.Of.Incident, Incident.City, Total.Amount.Of.Damages, Adjusted.Total.Damages, values) %>%
  arrange(-Adjusted.Total.Damages)

hazmat_region_adjusted_incidents %>%
    arrange(-Adjusted.Total.Damages) %>%
  
hazmat_region_adjusted_incidents %>%
  summarize(sum(Adjusted.Total.Damages))
```


#################################################################
Bar of US States by Incident Rate
#################################################################
- Use protracted data of all states from 2013-2023
- Use incidents
```{r}
hazmat_allstates_incidents %>%
  count(Incident.State) %>%
  arrange(-n)
```


#################################################################
Bar of Incidents by Mode
#################################################################
```{r}
incidents_by_mode <- hazmat_region_full_incidents %>%
  group_by(year, Mode.Of.Transportation) %>%
  summarise(count = n())

incidents_by_mode %>%
  pivot_wider(names_from = Mode.Of.Transportation, values_from = count)

```


#################################################################
Damages in the region Sankey
#################################################################
Average damages
```{r in the region}
hazmat_region_adjusted_incidents %>%
  group_by(Mode.Of.Transportation) %>%
  summarize(sum = sum(Adjusted.Total.Damages)) %>%
  arrange(-sum) %>%
  mutate(percentage = sum/sum(sum) *100)


hazmat_region_adjusted_incidents %>%  #2668.047
  ungroup() %>%
  filter(Mode.Of.Transportation == "Highway") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

hazmat_region_adjusted_incidents %>% #153747
  ungroup() %>%
  filter(Mode.Of.Transportation == "Rail") %>%
  summarise(mean_damages = mean(Adjusted.Total.Damages, na.rm = TRUE))

153747/2668.047

```

#################################################################
Urban/Rural
#################################################################

NCHS 
```{r}

hazmat_map_all <- hazmat_map_all %>%
  mutate(urban = case_when(
    X2013.code == 1 ~ "urban",
    X2013.code %in% c(2, 3, 4) ~ "suburban", 
    X2013.code %in% c(5, 6) ~ "rural"
  ))

#all incidents
hazmat_map_all %>%
  group_by(urban) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 

hazmat_map_all %>%
  filter(Serious.Incident.Ind == "Yes") %>%
  #filter(Transportation.Phase == 'In Transit') %>%
  group_by(urban) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 


#serious transit accidents
hazmat_map_all %>%
  filter(Serious.Incident.Ind == "Yes") %>%
  filter(Transportation.Phase == 'In Transit') %>%
  group_by(urban) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 

#about the same for highway and rail
hazmat_map_all %>%
  filter(Mode.Of.Transportation == "Rail" | Mode.Of.Transportation == "Highway") %>%
  #filter(Serious.Incident.Ind == "Yes") %>%
  #filter(Transportation.Phase == "In Transit") %>%
  group_by(Mode.Of.Transportation) %>%
  count(urban) %>%
  mutate(percentage = n / sum(n) * 100)

hazmat_map_all %>%
  filter(Mode.Of.Transportation == "Rail" | Mode.Of.Transportation == "Highway") %>%
  filter(Serious.Incident.Ind == "Yes") %>%
  filter(Transportation.Phase == "In Transit") %>%
  group_by(Mode.Of.Transportation) %>%
  count(urban) %>%
  mutate(percentage = n / sum(n) * 100)
```

Other urban/rural metrics, use for comparison purposes

RUCC
```{r omb ruc codes using counties}
omb_ruc_codes <- read.csv("/Users/koung/OneDrive - Gannett Company, Incorporated/Summer 2023 Projects/Hazmat_Incidents/final_data_product/ruralurbancodes2013.csv")

omb_ruc_codes <- omb_ruc_codes %>%
  mutate(countystate = paste(County_Name, State, sep = ", ")) %>%
  mutate(countystate = toupper(countystate)) %>%
  select(countystate, RUCC_2013)


hazmat_ten_full_omb_ruc <- left_join(hazmat_region_adjusted_incidents, omb_ruc_codes, by = "countystate")

hazmat_ten_full_omb_ruc <- hazmat_ten_full_omb_ruc %>%
  mutate(urban = case_when(
    RUCC_2013 == 1 ~ "urban",
    RUCC_2013 %in% c(2, 3) ~ "suburban", 
    RUCC_2013 %in% c(4, 5, 6, 7, 8, 9) ~ "rural"
  ))
  
hazmat_ten_full_omb_ruc %>%
  filter(!is.na(RUCC_2013)) %>%
  #filter(Serious.Incident.Ind == "Yes") %>%
  #filter(Transportation.Phase == 'In Transit') %>%
  group_by(urban) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 
```
HRSA
```{r}
hrsa_rural_counties <- read.csv("HRSA Rural Counties Designation.csv")

hrsa_county_codes <- hrsa_rural_counties %>%
  select(County, State, Metro) %>%
  mutate(countystate = paste(County, " COUNTY, ", State, sep = "")) 
  #%>% filter(Metro == 0)

hrsa_county_codes$countystate <- toupper(hrsa_county_codes$countystate)


hazmat_ten_full_hrsa <- left_join(hazmat_ten_full_incidents, hrsa_county_codes, by = "countystate")


hazmat_ten_full_hrsa <- hazmat_ten_full_hrsa %>%
  mutate(HRSA.Rural = ifelse(!is.na(County), "Yes", "No"))


hazmat_ten_full_hrsa %>%
  group_by(HRSA.Rural) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 

hazmat_ten_full_hrsa %>%
  filter(Serious.Incident.Ind == "Yes") %>%
  filter(Transportation.Phase == 'In Transit') %>%
  group_by(HRSA.Rural) %>%
  count() %>%
  ungroup() %>%
  mutate(percentage = (n / sum(n)) * 100) 
```

#################################################################
Incidents vs damages
#################################################################

```{r}
hazmat_region_adjusted_incidents %>%
  count(Mode.Of.Transportation)

```

```{r}
hazmat_region_adjusted_incidents %>%
  group_by(Mode.Of.Transportation) %>%
  summarize(sum = sum(Adjusted.Total.Damages)) %>%
  mutate(percentage = sum/sum(sum) *100)


hazmat_region_adjusted_incidents %>%
  filter(Transportation.Phase == "In Transit") %>%
  group_by(Mode.Of.Transportation) %>%
  summarize(sum = sum(Adjusted.Total.Damages)) %>%
  mutate(percentage = sum/sum(sum) *100)

```
